---
title: "Virginia Election Project"
author: "Joshua K. De La Torre"
format:
  html:
    self-contained: true
    code-fold: true
    code-tools: true
---

Introduction: This Code Walkthrough will use the joined_vacomparison dataset to analyze the electoral performance of presidential and gubernatorial candidates during the national elections. This walkthrough will be a step by step guide into the process of using datasets as potential sources for stories. This includes a comparison of total votes between candidates, % difference of votes, as well as steps to visualize them into tables and charts.

```{r setup, include=FALSE, warning=FALSE, message=FALSE}

#Load libraries
library(tidyverse)
library(janitor)
library(readxl) 
library(scales)
library(lubridate)
library(httr)
library(jsonlite)
library(kableExtra)
library(here)
options(scipen = 999)
options(stringsAsFactors = FALSE)
 

#load saved joined data file from previous script
joined_vacomparison <- readRDS(here("processed_data", "joined_vacomparison.rds"))

```

# Comparing Virginia Gov vs. Prez

```{r}

head(joined_vacomparison)

```

Include code to a new column to the dataset calculating the difference between the percentage of the vote Youngkin received and what Trump received.

```{r}

youngkinvstrump <- (joined_vacomparison) %>%
  group_by(trump_pct, pct_youngkin)

head(youngkinvstrump)

pct_diff_YT <- youngkinvstrump %>%
  group_by(trump_pct, pct_youngkin) %>%
  summarise(votemargin=pct_youngkin-trump_pct)

joined_vacomparison<- cbind(joined_vacomparison, pct_diff_YT$votemargin)
```

Use this new column to find the Top 5 counties with the highest differences between Younkin/Trump percent, and then create a well-labeled chart to visualize them.

```{r}
youngkincounties <- joined_vacomparison %>%
  group_by(locality)%>%
  arrange(-`pct_diff_YT$votemargin`)%>%
  head(5)%>%
  summarise(pctdifference=`pct_diff_YT$votemargin`)

youngkincounties %>%
  ggplot(aes(x = reorder(locality, pctdifference), y = pctdifference)) +
  geom_col(color = "#9ebcda", fill = "#9ebcda") +
  coord_flip() +
  scale_y_continuous(name = "PCT Difference", labels = scales::comma) +
  scale_x_discrete(name = "Top 5 Counties") +
  labs(title = "Vote Percentage Difference Between Youngkin & Trump") +
  theme_classic()

```

Create a chart that shows the Top 5 counties based on where Youngkin got the highest percentage of the vote

```{r}

youngkinwins <- joined_vacomparison %>%
  group_by(youngkin, trump) %>%
  arrange(-joined_vacomparison$youngkin+joined_vacomparison$trump)%>%
  head(5)%>%
  summarise(pcthigh=joined_vacomparison$youngkin+joined_vacomparison$trump)

youngkinhigh <- youngkinwins %>%
  mutate(
    pct_youngkin = janitor::round_half_up(youngkin / pcthigh * 100, 2), trump_pct = janitor::round_half_up(trump/ pcthigh *100, 2)
  ) %>%
  head(5)

youngkinhigh %>%
  ggplot(aes(x = reorder(youngkin, pcthigh), y = pcthigh)) +
  geom_col(color = "#9ebcda", fill = "#9ebcda") +
  coord_flip() +
  scale_y_continuous(name = "Vote %", labels = scales::comma) +
  scale_x_discrete(name = "Counties") +
  labs(title = "Top Youngkin Counties") +
  theme_classic()

```

Create a table that shows the Top 5 counties based on where McCauliffe got the highest percentage of the vote

```{r}
joined_vacomparison <- readRDS(here("processed_data", "joined_vacomparison.rds"))

McCauliffvstrump <- (joined_vacomparison) %>%
  group_by(pct_mcauliffe, trump_pct)

head(McCauliffvstrump)

pct_diff_MCT <- McCauliffvstrump %>%
  group_by(trump_pct, pct_mcauliffe) %>%
  summarise(votemargin=pct_mcauliffe-trump_pct)

joined_vacomparison<- cbind(joined_vacomparison, pct_diff_MCT$votemargin)


McCauliffecounties <- joined_vacomparison %>%
  group_by(locality)%>%
  arrange(-pct_diff_MCT$votemargin)%>%
  head(5)%>%
  summarise(pctdifference=pct_diff_MCT$votemargin)

McCauliffecounties %>%
  ggplot(aes(x = reorder(locality, pctdifference), y = pctdifference)) +
  geom_col(color = "#9ebcda", fill = "#9ebcda") +
  coord_flip() +
  scale_y_continuous(name = "PCT Difference", labels = scales::comma) +
  scale_x_discrete(name = "Top 5 Counties") +
  labs(title = "Vote Percentage Difference Between McCauliffe & Trump") +
  theme_classic()

```

Based on your own analysis, explain a finding of your choosing, include a narrative explaining it to the audience and then create either a chart or table displaying it.

```{r}
joined_vacomparison <- readRDS(here("processed_data", "joined_vacomparison.rds"))

governorrace <- (joined_vacomparison) %>%
  group_by(pct_mcauliffe, pct_youngkin, locality) %>%
  head(5) %>%
  summarise(votemargin = pct_youngkin-pct_mcauliffe)


head(governorrace)

pct_diff_MY <- governorrace %>%
  group_by(pct_mcauliffe, pct_youngkin, locality) %>%
  summarise(votemargin=pct_youngkin-pct_mcauliffe)

```

Of the top five counties with high margin of votes between candidates, three went to Youngkin who won the election. Both candidates won their respective counties with either 60-70% of the votes.

Based on your own analysis, explain a second finding of your choosing, include a narrative explaining it to the audience and then create either a chart or table displaying it.

```{r}
joined_vacomparison <- readRDS(here("processed_data", "joined_vacomparison.rds"))

governorvotes <- (joined_vacomparison) %>%
  group_by(youngkin) %>%
  head(5)

governors<- governorvotes %>%
  group_by(youngkin, mcauliffe)%>%
  head(5)%>%
  summarise()

ggplot(governorvotes, aes(x = governors$youngkin, y = locality)) +
  geom_col()+
  scale_fill_manual(values = c("#468189", "#9DBEBB"))

```

These Graphs visualize the support both candidates receieved in the conties with most votes. Youngkin had a huge lead in Amelia and Alleghany with moderate support from Alexandria and Albemarble. While McCauliffe still dominated Alexandria and Albemarble, he had even fewer support in Amelia and Alleghany. Youngkin would do well even in counties he lost, whole McCauliffe barely registers in counties he lost. Patterns like these likely led to Youngkin's election win.

```{r}
ggplot(governorvotes, aes(x = governorvotes$locality, y = governorvotes$mcauliffe)) +
  geom_col()+
  coord_flip()+
  scale_fill_manual(values = c("#468189", "#9DBEBB"))
```

Based on your own analysis, explain a third finding of your choosing, include a narrative explaining it to the audience and then create either a chart or table displaying it.

```{r}
joined_vacomparison <- readRDS(here("processed_data", "joined_vacomparison.rds"))

presZ_race <- (joined_vacomparison) %>%
  group_by(biden_pct, trump_pct) %>%
  head(5)

govs_race <- (joined_vacomparison)

govs_race <- (joined_vacomparison) %>%
  group_by(govs_race$pct_youngkin, govs_race$pct_mcauliffe) %>%
  head(5)


pct_diff_elec <- merge(govs_race, presZ_race) %>%
  group_by(locality) %>%
  summarise(votecount=mean(biden_pct-trump_pct), mean(pct_youngkin-pct_mcauliffe))

head(pct_diff_elec)

```

In This table is the mean difference or gap of votes between presidential and gubernatorial candidates. The difference in votes seems to be slightly higher in the gubenatorial race compared to the presidential race.

```{r}
pct_diff_elec %>%
  ggplot(aes(x= reorder(locality, votecount), y = votecount ))+
  geom_col(color = "#9ebcda", fill = "#9ebcda")+
  coord_flip()+
  scale_y_continuous(name = "Votecount", labels = scales::comma) +
  scale_x_discrete(name = "County") +
  labs(title = "Mean Total votes between Governor and Presidential Candidates in Virginia", 
       subtitle = "2020") +
  theme_classic()
```
